Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-09 WholePlotTrajectories.Rmd
graphs

The taxonomic diversity was reported through species richness and evenness, *i.e* the Hill number translation of the Simpson index [@Hill1973].
These indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0 and 2 order of diversity (q), recomended for diversity studies [@Patil1982;@Tothmeresz1995;@Marcon2015]. 
The functional diversity was reported using the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on species functional traits.

@@

```{r DivTaxo, out.width = '100%', fig.asp = 1.5/3, echo=FALSE, fig.cap="Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic (ref:Bolda) richness, (ref:Boldb), taxonomic evenness and (ref:Boldc) functional diversity. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals ", fig.env="figure*"}
load("DB/ReplacementTraj_ForGraphs")
load("DB/FunctionalTraj_ForGraphs")

par(mfrow=c(1,3),mar=c(5,2,4,2),oma=c(1,1.5,1,1),no.readonly=TRUE)
TaxoTraj(CompleteTaxo)
plotDiv(CompleteFun,remove=TRUE)
mtext("(c) Functional diversity",side=3,adj=0,line=1)
mtext("Years since disturbance",side=1,line=2.2,adj=1)
legend("right",inset=c(-0.34,0),xpd=NA,legend=c("T0","T1","T2","T3"),col=ColorsTr,lwd=2.5,bty="n",title="Treatment")

@@

The plot 7 from treatment 1 displayed a constantly outlying functional diversity and was removed from the graphical representation for better readability (see appendix for full graphs).
In undisturbed plots the functional diversity remained stable along the 30 years. In disturbed plots, trajectories depend on the disturbnace intensity with, for low intensity, a low but long-lasting increase up to amximum reached after 20-25 years and, for, high intensity, a fast but short increase followed, after 10 years, by a slow decrease towards the inital values.

The second-degree polynomial regressions between (i) the %AGB loss and (ii) taxonomic richness, evenness and functional diversity after 10, 20 and 30 years best predicted the hump-shaped curve of the disturbance impact along the disturbance intensity gradient \@ref(fig:IDHplot).
Along the disturbance intensity gradient the disturbance impact was more markedly hump-shaped for the taxonomic richness than for the taxonomic evenness and the functional diversity, the latest being almost linear.
The regression model better predicted the functional diversity ($0.60<R^2_{Functional}<0.81$) than the taxonomic richness and evenness ($0.21<R^2_{Richness}<0.4$, and $-0.15<R^2_{Evenness}<0.43$ respectively)

```{r IDHplot, out.width = '100%', echo=FALSE, fig.asp = 1.5/3,fig.cap="Relationship between the initial \\%AGB loss and community taxonmic richness (ref:Bolda), taxonomic evenness (ref:Boldb) and functional diversity (ref:Boldc) at 10, 20 and 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals ", fig.env="figure*"}
load("DB/RichnessIDH");load("DB/SimpsonIDH");load("DB/RaoIDH");load("DB/LostAGB")

par(mfrow=c(1,3),mar=c(5,2,4,2),oma=c(1,1.5,1,1),no.readonly=TRUE)
plotIDH(Data=Rich,AgbLoss=AGBloss)
mtext("(a) Taxonomic Richness",line=1,side=3,adj=0,cex=0.8)
plotIDH(Data=Sim,AgbLoss=AGBloss)
mtext("(b) Taxonomic Evenness",line=1,side=3,adj=0,cex=0.8)
plotIDH(Data=Rao,AgbLoss=AGBloss)
mtext("(c) Functional diversity",line=1,side=3,adj=0,cex=0.8)
mtext("initial AGB loss (%)",side=1,line=2.2,adj=1)
legend("right",inset=c(-0.34,0),xpd=NA,legend=c("10","20","30"),col=colyear,lwd=2.5,bty="n",title="Years")

@@

Disturbance clearly enhanced community taxonomic richness until an intensity threshold (20-25% of initial AGB loss), while the taxonomic evenness was somewhat decoupled from the disturbance intensity, as already observed in the Guiana Shield [@Baraloto2012a] and in Bornean tropical forests [@Cannon1998].
The IDH well predicted the response of taxonomic richness to disturbance, and to a lesser extent the taxonomic evenness.
Post-disturbance communities combine pre-disturbance surviving trees and recruited trees, either from the remaining pre-disturbance community or from a disturbance-specific community.
In the Guiana Shield, the pool of true pioneer species specifically recruited after disturbance is restricted to a few common genera (e.g. Cecropia spp., Vismia spp.) 
Below the intensity threshold the taxonomic richness of the pre-disturbance community is maintained and the recruitment of pioneers, previously infrequent or absent, increases both taxonomic richness and evenness.
The disturbance intensity determined the relative importance of pioneers recruited, so the taxonomic richness and evennes increase all the more so that disturbance is intense.
However, disturbance in the same time decreases the taxonomic richness of the pre-disturbance community, and beyond the intensity threshold the decrease exceeds the taxonomic enrichment from pioneers.
The taxonomic evenness for its part is less sensitive to the loss of rare species and the disturbance impact is mitigated as pioneers balance the hyper-dominance of undisturbed tropical forests.

Conversely the IDH was disproved for the functional diversity.
Pioneer species are functionally highly different, and whenever the disturbance intensity their increased abundance suffice to increase the community functional diversity and evenness.

Following disturbance the taxonomic richness droped at first place independently of the disturbance intensity, but then increased through a species gain that was dermined by the disturbance intensity.
Below the intensity threshold the establishment of long-lived pioneers significantly enhanced both taxonomic and functional richness and evenness in the long-term.
Beyond the intensity threshold the establishment of short-lived pioneers rapidly filling the environmental space made available did not increased the taxonomic richness and entailed a decrease in the functional diversity after 10 years.
Likely, after intense disturbance the long-lived pioneers will succeed and post-disturbance trajectories will match those of intermediate disturbance, which explains the decoupling in time between high and low disturbance intensity trajectories.

Species first established after disturbance remained in the community but became less dominant, replaced the abundant pre-disturbance species, but the recovery of rare species was hampered in the long-term [@Hubbell2001;@Chave2004].

## The functional redundancy, key of the taxonomic resilience

The recovery of the most infrequent species followed the lottery recruitment rules [@Busing2002] and was hampered by the increasing competition among species, implying long term recovery [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Following disturbance the space and resources made available corresponded to a rapid decrease of the functional redundancy within the initial functional space.
The high light, space and nutrient niches were rapidly filled by species, mainly the most dominant and frequent ones.
Thereafter competitive exclusion emerged following the filling of the niches and limited the recruitment of infrequent species [@Busing2002].
The recovery of communities taxonomic richness and evenness relied upon the recovery of the initial functional redudancy.
Although underway 30 years after disturbance the recovery remained unachieved and after 30 years the recovery it was underway but remained unachieved.
This alteration of the functional redundancy meant a lower resilience of pre-disturbance communities and higher chances to see the persistence of disturbance-specific communities, with less species and more pioneers [@Haddad2008; @Burslem2000; @Martin2013].
Besides the slowed recovery of rare species increased the risks to loose cornerstone specie, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].
Apart from the functional characteristics considered here, infrequent species mighthave unique functions i nthe ecosystem or be a key for some fauna.

@@

Our study revealed communities cyclic recovery after disturbance allowing the resilience of their functioning and taxonomic composition with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for  20 years after disturbance through the enrichment of the communities with pioneers and light-demanding species,in accordance with the IDH.
The IDH, though, poorly predicted the disturbance impact on communities taxonomic richness and evenness that were blurried by the emergence of competitive exclusion along time.
The resilience of tropical forests proved consistent although spread over several decades. 
Still, the disturbance impact on communities redundancy cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Gourlet-Fleury2005].
As the trajectories highlighted the recruitment processes proved central for communities response to disturbance and closer focus on demographical drivers of communities response would clarify the fate of the future forests.

>>>>

The taxonomic diversity was reported through species richness and evenness, *i.e* the Hill number translation of the Simpson index [@Hill1973].
These indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0 and 2 order of diversity (q), recomended for diversity studies [@Patil1982;@Tothmeresz1995;@Marcon2015]. 
The functional diversity was reported using the functional richness and evenness, *i.e* Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on species functional traits.

@@

```{r DivTaxo, out.width = '100%', fig.asp = 1.5/3, echo=FALSE, fig.cap="Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic (ref:Bolda) richness, (ref:Boldb), taxonomic evenness, (ref:Boldc) functional richness, and (ref:Boldd) functional evenness. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals ", fig.env="figure*"}
load("DB/ReplacementTraj_ForGraphs")
load("DB/FunctionalTraj_ForGraphs")
load("DB/FunctionalRichnessTraj_ForGraphs")

par(mfrow=c(1,4),mar=c(5,2,4,2),oma=c(1,1.7,1,2),no.readonly=TRUE)
TaxoTraj(CompleteTaxo)
plotDiv(CompleteRichFun,remove=TRUE)
mtext("(c) Functional Richness",side=3,adj=0,line=1,cex=0.75)
plotDiv(CompleteFun,remove=TRUE)
mtext("(d) Functional Evenness",side=3,adj=0,line=1,cex=0.75)
mtext("Years since disturbance",side=1,line=2.2,adj=1)
legend("right",inset=c(-0.34,0),xpd=NA,legend=c("T0","T1","T2","T3"),col=ColorsTr,lwd=2.5,bty="n",title="Treatment")

@@

The plot 7 from treatment 1 displayed constantly outlying functional richness and evenness and was removed from the graphical representation for better readability.
In undisturbed plots both functional richness and evenness remained stable along the 30 years. 
In disturbed plots, richness and evenness trajectories depended on the disturbance intensity with their maximum positively correlated to %AGB loss $\rho_{spearman}^{Richness}=0.76$ and $\rho_{spearman}^{Evenness}=0.60$.
Functional richness and evenness displayed for low disturbance intensity a low but long-lasting increase up to maximum reached after 20-25 years and, and for high intensity, a fast but short increase followed after 10 years by a slow decrease towards the inital values.

The second-degree polynomial regressions between (i) the %AGB loss and (ii) taxonomic and functional richness and evenness after 10, 20 and 30 years best predicted the hump-shaped curve of the disturbance impact along the disturbance intensity gradient \@ref(fig:IDHplot).
Along the disturbance intensity gradient the disturbance impact was more markedly hump-shaped for the taxonomic richness than for the taxonomic evenness and the functional diversity, the latest being almost linear.
The regression model better predicted the functional richness and evenness ($0.55<R^2_{Functional Richness}<0.72$, and $0.60<R^2_{Functional Evenness}<0.81$) than the taxonomic richness and evenness ($0.21<R^2_{Taxonomic Richness}<0.4$, and $-0.15<R^2_{Taxonomic Evenness}<0.43$ respectively)

```{r IDHplot, out.width = '100%', echo=FALSE, fig.asp = 1.5/3,fig.cap="Relationship between the initial \\%AGB loss and community taxonmic richness (ref:Bolda), taxonomic evenness (ref:Boldb), functional richness (ref:Boldc),and functional evenness (ref:Boldd) at 10, 20 and 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals ", fig.env="figure*"}
load("DB/RichnessIDH");load("DB/SimpsonIDH");load("DB/RaoIDH");load("DB/LostAGB");load("DB/FunRichIDH")

par(mfrow=c(1,4),mar=c(5,2,4,2),oma=c(1,1.5,1,1),no.readonly=TRUE)
plotIDH(Data=Rich,AgbLoss=AGBloss)
mtext("(a) Taxonomic Richness",line=1,side=3,adj=0,cex=0.75)
plotIDH(Data=Sim,AgbLoss=AGBloss)
mtext("(b) Taxonomic Evenness",line=1,side=3,adj=0,cex=0.75)
plotIDH(Data=Rao,AgbLoss=AGBloss)
mtext("(c) Functional richness",line=1,side=3,adj=0,cex=0.75)
plotIDH(Data=FunRich,AgbLoss=AGBloss)
mtext("(d) Functional evenness",line=1,side=3,adj=0,cex=0.75)
mtext("initial AGB loss (%)",side=1,line=2.2,adj=1)
legend("right",inset=c(-0.34,0),xpd=NA,legend=c("10","20","30"),col=colyear,lwd=2.5,bty="n",title="Years")

@@

Disturbance appeared to enhance community taxonomic richness until an intensity threshold (20-25% of initial AGB loss), while the taxonomic evenness was somewhat decoupled from the disturbance intensity as already observed in the Guiana Shield [@Baraloto2012a] and in Bornean tropical forests [@Cannon1998].
The IDH well predicted the disturbance impact on community taxonomic richness and to some extent on taxonomic evenness.
The disturbance intensity determines the balance in the community between pre-disturbance surviving trees and the trees recruited afterward.
In the Guiana Shield, the pool of true pioneer species specifically recruited after disturbance is restricted to a few common genera (e.g. Cecropia spp., Vismia spp.).
Below the intensity threshold the size of the surviving community maintains its pre-disturbance taxonomic richness and the recruitment of pioneers, infrequent or absent before disturbance, increases both community taxonomic richness and evenness.
Beyond the intensity threshold the taxonomic diversity of the surviving trees decreases and is not offset by the taxonomic enrichment of pioneers.
The disturbance impact was similar but slighter for community taxonomic evenness, that is less sensitive to the loss of rare species and rather represents the increased dominance of pionneers that balanced the usual hyper-dominance of tropical forest.
While below the intensity threshold pioneers species increase the diversity of dominant or common species, beyond the threshold they become in turn highly dominant.

Conversely the IDH was disproved to predict the dsiturbance impact on community functional diversity.
Pioneer species were functionally highly different from the pre-disturbance community composition, so their recruitment increased the community functional richness and eveness whenever its intensity.

Along time, the post-disturbance taxonomic richness trajectories showed first a similar drop following the species loss during disturbance, and then a species gain determined by the disturbance intensity.
Until an intensity threshold the gain in taxonomic richness increased with the disturbance intensity through the establishment of long-lived pioneers enhancing community taxonomic richness and evenness in the long term.
These long-lived pionneers were functionally quite different from the functional composition and their establishment entailed a progressive and long-lasting increase of the functional diversity.
Beyond the intensity threshold a few short-lived pioneers occupied the vacated environmental space and prevented further species establishment. 
These short-lived pioneers were functionally very different from the pre-disturbance community and rapidly and significantly increased the functional diversity.
Already after 10 years, though, they started to decline and the functional diversity started to decrease.
Likely the establishment of long-lasting pioneers will follow and we expect the taxonomic and functional trajectories to catch up with those observed after intermediate disturbance.

## The functional redundancy, key of the community resilience

For 15 years the species loss during disturbance, determined by the disturbance intensity, commensurately decreased the functional redundancy within the pre-disturbance functional space.
At first place the pioneers recruited, functionally different from the pre-disturbance functional composition did not compensate for the reundancy decrease.
Progressively though, first establish species were replaced by more competitive long-lived pionneers or late-successional species resembling more the pre-disturbance functional composition and restoring the functional redundancy.
This replacement was stochastic and followed the lottery recruitment rules, that ease the recovery for the first recruited species but then increasingly hamper it with the emergence of interspecific competition due to increasing functional redundancy [@Busing2002]. 
Along time the recovery of infrequent species is increasingly difficult, and the recovery of the functional redundancy that was just initiated after 30 years in some communities is difficult to estimate [@Trenbath1999;@Elmqvist2003;@Diaz2005].

The long-term impact of disturbance on community functional redundancy meant a lower resilience of pre-disturbance communities and higher chances to see the persistence of disturbance-specific species at the expense of late-successional ones [@Haddad2008; @Burslem2000; @Martin2013].
Besides the long-term recovery of infrequent species increases the risks to loose cornerstone species, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].
Apart from the functional characteristics considered here, infrequent species might indeed have unique functions in the ecosystem or be a key for some fauna.

@@

[Our study revealed communities cyclic recovery after disturbance allowing the resilience of their functioning and taxonomic composition with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for  20 years after disturbance through the enrichment of the communities with pioneers and light-demanding species,in accordance with the IDH.
The IDH, though, poorly predicted the disturbance impact on communities taxonomic richness and evenness that were blurried by the emergence of competitive exclusion along time.]
The resilience of tropical forests proved consistent although several decades-long. 
Still, the disturbance impact on communities redundancy cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Gourlet-Fleury2005].
As the trajectories highlighted the recruitment processes proved central for communities response to disturbance and closer focus on demographical drivers of communities response would clarify the fate of the future forests.

### 2018-07-09 WholePlotTrajectories.tex
graphs

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the Rao index of quadratic entropy which combines
species abundance distribution and average pairwise dissimilarity based
on species functional traits.

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic \textbf{(a)} richness, \textbf{(b)}, taxonomic evenness and \textbf{(c)} functional diversity. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

The plot 7 from treatment 1 displayed a constantly outlying functional
diversity and was removed from the graphical representation for better
readability (see appendix for full graphs). In undisturbed plots the
functional diversity remained stable along the 30 years. In disturbed
plots, trajectories depend on the disturbnace intensity with, for low
intensity, a low but long-lasting increase up to amximum reached after
20-25 years and, for, high intensity, a fast but short increase
followed, after 10 years, by a slow decrease towards the inital values.

The second-degree polynomial regressions between (i) the \%AGB loss and
(ii) taxonomic richness, evenness and functional diversity after 10, 20
and 30 years best predicted the hump-shaped curve of the disturbance
impact along the disturbance intensity gradient \ref{fig:IDHplot}. Along
the disturbance intensity gradient the disturbance impact was more
markedly hump-shaped for the taxonomic richness than for the taxonomic
evenness and the functional diversity, the latest being almost linear.
The regression model better predicted the functional diversity
(\(0.60<R^2_{Functional}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Richness}<0.4\), and \(-0.15<R^2_{Evenness}<0.43\)
respectively)

@@

\caption{Relationship between the initial \%AGB loss and community taxonmic richness \textbf{(a)}, taxonomic evenness \textbf{(b)} and functional diversity \textbf{(c)} at 10, 20 and 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:IDHplot}
\end{figure*}

@@

Disturbance clearly enhanced community taxonomic richness until an
intensity threshold (20-25\% of initial AGB loss), while the taxonomic
evenness was somewhat decoupled from the disturbance intensity, as
already observed in the Guiana Shield \citep{Baraloto2012a} and in
Bornean tropical forests \citep{Cannon1998}. The IDH well predicted the
response of taxonomic richness to disturbance, and to a lesser extent
the taxonomic evenness. Post-disturbance communities combine
pre-disturbance surviving trees and recruited trees, either from the
remaining pre-disturbance community or from a disturbance-specific
community. In the Guiana Shield, the pool of true pioneer species
specifically recruited after disturbance is restricted to a few common
genera (e.g.~Cecropia spp., Vismia spp.) Below the intensity threshold
the taxonomic richness of the pre-disturbance community is maintained
and the recruitment of pioneers, previously infrequent or absent,
increases both taxonomic richness and evenness. The disturbance
intensity determined the relative importance of pioneers recruited, so
the taxonomic richness and evennes increase all the more so that
disturbance is intense. However, disturbance in the same time decreases
the taxonomic richness of the pre-disturbance community, and beyond the
intensity threshold the decrease exceeds the taxonomic enrichment from
pioneers. The taxonomic evenness for its part is less sensitive to the
loss of rare species and the disturbance impact is mitigated as pioneers
balance the hyper-dominance of undisturbed tropical forests. Conversely
the IDH was disproved for the functional diversity. The pioneer species
were functionally highly different from the pre-disturbance community
composition, so their recruitment increased the community functional
richness and eveness after disturbance whenever its intensity.

Community post-disturbance trajectories showed a first drop of the
taxonomic richness, similar among communities, followed by a species
gain determined by the disturbance intensity. Until an intensity
threshold the gain in taxonomic richness increased with the disturbance
intensity. This corresponded to the establishment of long-lived pioneers
enhancing community taxonomic richness and evenness in the long term.
Beyond the intensity threshold a few short-lived pioneers occupied the
vacated environmental space, preventing the establishment of further
species. The establishment of short-lived pioneers, functionally very
different from the pre-disturbance community, rapidly and significantly
increased the functional diversity but it started decreasing after 10
years. Below the disturbance intensity threshold on the other hand the
establishment of long-lived pionneers entailed a later and lower but
long-lasting increase of the functional diversity. Likely the
establishment of long-lasting pioneers will follow the decline of
short-live pionneers and we expect the taxonomic and functoinal
trajectories to catch up with those observed after intermediate
disturbance.

\subsection{The functional redundancy, key of the taxonomic
resilience}\label{the-functional-redundancy-key-of-the-taxonomic-resilience}

The functional redundancy within the pre-disturbance functional space
decreased for 15 years because of the loss of species, and was not
compensated by the functionally different pionneers. The depth of the
redundancy decrease relied on the disturbance intensity that enhanced
the initial decrease and the functional differences of pionneers from
the pre-disturance composition. After 15 years the first established
pioneers pionneers started to be replaced by more competitive long-lived
pionneers or late-successional species resembling more the
pre-disturbance functional composition and hence increasing the
functional redundancy. This replacement though is stochastic and follows
the lottery recruitment so because of increasing competition it is
increasingly difficult recover pre-disturbance species,specifically the
most infrequent ones, as the functional redundancy recovers
\citep{Busing2002}. The recovery time proved long as for some community
it was just initiated after 30 years but it is thus difficult to
estimate \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of pre-disturbance communities and higher
chances to see the persistence of disturbance-specific communities, with
lower taxonomic richness and more pioneers
\citep{Haddad2008, Burslem2000, Martin2013}. Besides the slowed recovery
of rare species increased the risks to loose cornerstone specie, with
unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species mighthave
unique functions i nthe ecosystem or be a key for some fauna.

@@

Our study revealed communities cyclic recovery after disturbance
allowing the resilience of their functioning and taxonomic composition
with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for 20 years after
disturbance through the enrichment of the communities with pioneers and
light-demanding species,in accordance with the IDH. The IDH, though,
poorly predicted the disturbance impact on communities taxonomic
richness and evenness that were blurried by the emergence of competitive
exclusion along time. The resilience of tropical forests proved
consistent although spread over several decades. Still, the disturbance
impact on communities redundancy cautioned against the risks of
infrequent species loss and the persistence of disturbance-specific
communities \citep{Gourlet-Fleury2005}. As the trajectories highlighted
the recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.

>>>>

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the functional richness and evenness, \emph{i.e} Rao
index of quadratic entropy which combines species abundance distribution
and average pairwise dissimilarity based on species functional traits.

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic \textbf{(a)} richness, \textbf{(b)}, taxonomic evenness, \textbf{(c)} functional richness, and \textbf{(d)} functional evenness. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

The plot 7 from treatment 1 displayed constantly outlying functional
richness and evenness and was removed from the graphical representation
for better readability. In undisturbed plots both functional richness
and evenness remained stable along the 30 years. In disturbed plots,
richness and evenness trajectories depended on the disturbance intensity
with their maximum positively correlated to \%AGB loss
\(\rho_{spearman}^{Richness}=0.76\) and
\(\rho_{spearman}^{Evenness}=0.60\). Functional richness and evenness
displayed for low disturbance intensity a low but long-lasting increase
up to maximum reached after 20-25 years and, and for high intensity, a
fast but short increase followed after 10 years by a slow decrease
towards the inital values.

The second-degree polynomial regressions between (i) the \%AGB loss and
(ii) taxonomic and functional richness and evenness after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient \ref{fig:IDHplot}. Along the
disturbance intensity gradient the disturbance impact was more markedly
hump-shaped for the taxonomic richness than for the taxonomic evenness
and the functional diversity, the latest being almost linear. The
regression model better predicted the functional richness and evenness
(\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Evenness}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Evenness}<0.43\) respectively)

@@

\caption{Relationship between the initial \%AGB loss and community taxonmic richness \textbf{(a)}, taxonomic evenness \textbf{(b)}, functional richness \textbf{(c)},and functional evenness \textbf{(d)} at 10, 20 and 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:IDHplot}
\end{figure*}

@@

Disturbance appeared to enhance community taxonomic richness until an
intensity threshold (20-25\% of initial AGB loss), while the taxonomic
evenness was somewhat decoupled from the disturbance intensity as
already observed in the Guiana Shield \citep{Baraloto2012a} and in
Bornean tropical forests \citep{Cannon1998}. The IDH well predicted the
disturbance impact on community taxonomic richness and to some extent on
taxonomic evenness. The disturbance intensity determines the balance in
the community between pre-disturbance surviving trees and the trees
recruited afterward. In the Guiana Shield, the pool of true pioneer
species specifically recruited after disturbance is restricted to a few
common genera (e.g.~Cecropia spp., Vismia spp.). Below the intensity
threshold the size of the surviving community maintains its
pre-disturbance taxonomic richness and the recruitment of pioneers,
infrequent or absent before disturbance, increases both community
taxonomic richness and evenness. Beyond the intensity threshold the
taxonomic diversity of the surviving trees decreases and is not offset
by the taxonomic enrichment of pioneers. The disturbance impact was
similar but slighter for community taxonomic evenness, that is less
sensitive to the loss of rare species and rather represents the
increased dominance of pionneers that balanced the usual hyper-dominance
of tropical forest. While below the intensity threshold pioneers species
increase the diversity of dominant or common species, beyond the
threshold they become in turn highly dominant.

Conversely the IDH was disproved to predict the dsiturbance impact on
community functional diversity. Pioneer species were functionally highly
different from the pre-disturbance community composition, so their
recruitment increased the community functional richness and eveness
whenever its intensity.

Along time, the post-disturbance taxonomic richness trajectories showed
first a similar drop following the species loss during disturbance, and
then a species gain determined by the disturbance intensity. Until an
intensity threshold the gain in taxonomic richness increased with the
disturbance intensity through the establishment of long-lived pioneers
enhancing community taxonomic richness and evenness in the long term.
These long-lived pionneers were functionally quite different from the
functional composition and their establishment entailed a progressive
and long-lasting increase of the functional diversity. Beyond the
intensity threshold a few short-lived pioneers occupied the vacated
environmental space and prevented further species establishment. These
short-lived pioneers were functionally very different from the
pre-disturbance community and rapidly and significantly increased the
functional diversity. Already after 10 years, though, they started to
decline and the functional diversity started to decrease. Likely the
establishment of long-lasting pioneers will follow and we expect the
taxonomic and functional trajectories to catch up with those observed
after intermediate disturbance.

\subsection{The functional redundancy, key of the community
resilience}\label{the-functional-redundancy-key-of-the-community-resilience}

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. At first place
the pioneers recruited, functionally different from the pre-disturbance
functional composition did not compensate for the reundancy decrease.
Progressively though, first establish species were replaced by more
competitive long-lived pionneers or late-successional species resembling
more the pre-disturbance functional composition and restoring the
functional redundancy. This replacement was stochastic and followed the
lottery recruitment rules, that ease the recovery for the first
recruited species but then increasingly hamper it with the emergence of
interspecific competition due to increasing functional redundancy
\citep{Busing2002}. Along time the recovery of infrequent species is
increasingly difficult, and the recovery of the functional redundancy
that was just initiated after 30 years in some communities is difficult
to estimate \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of pre-disturbance communities and higher
chances to see the persistence of disturbance-specific species at the
expense of late-successional ones
\citep{Haddad2008, Burslem2000, Martin2013}. Besides the long-term
recovery of infrequent species increases the risks to loose cornerstone
species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}. Apart from the
functional characteristics considered here, infrequent species might
indeed have unique functions in the ecosystem or be a key for some
fauna.

@@

{[}Our study revealed communities cyclic recovery after disturbance
allowing the resilience of their functioning and taxonomic composition
with the maintenance of initial differences among communities.
Communities functional evenness was enhanced for 20 years after
disturbance through the enrichment of the communities with pioneers and
light-demanding species,in accordance with the IDH. The IDH, though,
poorly predicted the disturbance impact on communities taxonomic
richness and evenness that were blurried by the emergence of competitive
exclusion along time.{]} The resilience of tropical forests proved
consistent although several decades-long. Still, the disturbance impact
on communities redundancy cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Gourlet-Fleury2005}. As the trajectories highlighted the
recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.
